Very Large Integers and polynomial inner products

### 2013-01-14 doc/paper_vli/vli.tex
morning reading paper VLI

%\date{}                                           % Activate to display a given date or no dater
\begin{document}
\maketitle
%\section{}
%\subsection{}
%\vspace{-0.65cm}
\begin{abstract}
The main purpose of the  Very Large Integer (VLI/C++/ASM) library is to perform large integer arithmetic and  inner products of polynomials. Based on meta-programming of assembly language,
and  hybrid (CPU/GPU) mode, it performs inner products of dense and triangular polynomials  of 1 to 4 variables  where the coefficients are represented  by large integers of 128 to 512 bits. 
In this paper, we present the basics of the inner products of polynomials, and  the implementation of the library for the CPU and the GPU.  All our algorithms are based on brute force attacks, because the problem is too small for a FFT approach. We save a factor up to 10 on CPU, and 60 on GPU compared to a GMP solution,
and attain 67 \% of the peak performance of the GPU.
\end{abstract}
\section{Introduction}
% TODO references LINBOX, NTL

The Very Large Integer library performs  standards  operations for large integer arithmetic and polynomial. This library also provides optimized implementations of inner products  for GPU.
This library is composed of two distinct modules,  but enough closed  to work together by the mechanism of template instantiation. VLI is a fix size library from 128 to 512 bits,
as the polynomial module  which is limited for dense and triangular polynomials  up to 4 variables. The core of VLI is build under ASM language to provide the best performance.  Every arithmetic operations
are executed by a specific kernel generated specifically during the compilation by an ASM generator. Moreover, VLI allows multithreading and hybrid  GPU usage for inner products of polynomials.
Multithreading is build on OpenMP library and GPU module under CUDA. As the CPU, the GPU is a fixe size library, and ASM kernels are also determinated specifically during the compilation.

In large arithmetic, the standard library is GMP\footnote{GMP is the standard library for long arithmetic, gmplib.org}. GMP is polyvalent, nevertheless this polyvalence is attained by the usage of  dynamic memory.
It is well know dynamic allocation is time consuming, and it can become a bottle neck for small operations which are in the working range.  
Alternatively for polynomial arithmetic, a few libraries like  LINBOX or NTL  provide a solution but  they use GMP when needed. 
Although hardware accelerators were introduced  in HPC a few years ago, as the GPUs  by NVIDIA. Only a few libraries have been adapted 
A prominent example of such an adaption are the MAGMA libraries, which port the well known BLAS/LAPACK packages to GPU accelerators.
The polynomials libraries, however,  have not done the transition to these new hardwares yet.
During the last few years, the polynomial multiplication was successfully implemented on this new hardware
\cite{Govindaraju2008, Emeliyanenko:2009, Moreno2010}. All these works focus on large polynomials, using a FFT approach.

In this work, we are going to highlight the speedup given by a tiny ASM library without any dynamic memory, suggest an alternative for small size integer. 
And propose an efficient implementation of inner product on GPU with the latest CUDA PTX features.

@@

The library performs standards operations for polynomials (up to four variables) and optimized solvers for  the inner products of vectors of polynomials (presented in the next section).
The polynomials are defined like  e.g. our variables
\begin{eqnarray}
P_a(x,y,z,w) & = & \sum_{i=0}^n \sum_{j=0}^n  \sum_{k=0}^n \sum_{l=0}^n  a_{ijkl} x^i y^j z^k w^l , \,\,\,\,\,\, \text{or} \\
                   & = & \sum_{i=0}^n \sum_{j=0}^{(n-i)} \sum_{k=0}^{(n-i-j)} \sum_{l=0}^{(n-i-j-k)}  a_{ijkl} x^i y^j z^k w^l  .
\end{eqnarray}
where  $x,\,y,\,z$ and $w$ are symbolic variables,  $i,\,j,\,k$ and $l$  the indices, $a$  the coefficients, and $n$ the order of the polynomial.
The polynomials of the form  (1) and (2) are named dense and triangular respectively. Consider now two vectors of polynomials of the  same type with $m$ 
entries  $\boldsymbol{v}_1$ and  $\boldsymbol{v}_2$. The inner products  consist of the product 
of all entries  associated with a global reduction. 
 \begin{eqnarray}
 P_c = \sum_{i=0}^m \boldsymbol{v}_{1_i} \times  \boldsymbol{v}_{2_i}  \label{poly1} \label{InnerProduct}
\end{eqnarray}
where $\times$ indicates the multiplication between two polynomials. The multiplications use usual rules of polynomials multiplication. 
A final coefficient may be the sum of several coefficients (which we call contributions in the following passage), resulting from the sum of several multiplications. These coefficients are the cross terms. 
The cross terms are not a problem for the CPU version, as a polynomial multiplication is done by a single thread, but they are the bottleneck for the GPU version.
The number of coefficient for a dense and triangular polynomial is given by $(n+1)^{\kappa}$ and  $(\kappa+n-1)! / ( (\kappa-1)! n! )$ respectively, where $\kappa$ is the number of variables.

@@

\section{VLI  CPU library}

@@

We coded the kernels using the GNU inline assembly.
Generating by hand all these  assembly strings (kernels)  can be painful and error-prone\footnote{The library has near 10000 lines of assembly in total for x86-64, Power64, and PTX}.
We designed a assembly generator. It creates all versions of a kernel during the compilation.   
It is built on Boost preprocessor package. We can thus create and manipulate the ASM strings during the compilation, 
and generate a specifically tuned kernel for every large integer. We give a small example of a long addition with the generator into the table \ref{ASMGENERATOR}.

@@

\subsection{Memory layout and SIMD possibility}

@@

We have choose the SoA for the two following reasons :

@@

The disagreement  of this choice, the SoA order does not facilitate the usage of the SIMD instruction such as the AVX instructions of the Sandy Bridge x86-64 architecture.
Applying SIMD to the AoS, for an addition or a subtraction is a nonsense because there is no specific instruction to propagate carry bit between entries of the SIMD register.
It could be  apply  on multiplications for integer  larger than 1024-bit \cite{SIMD}, unfortunately, it does not correspond to our working range. Consequently, we do not 
use the multimedia instruction into the ASM kernels.

@@

The GPU algorithm of the inner product is divided into two parts. A preliminary step on the CPU,  where we prepare the calculation blue print of the  contribution. And the calculation itself on the GPU

@@

\item  First, we create a dummy resulting polynomial  represented by the indices and the contributions. They are sorted by descending order of the number of contributions
and grouped by 32 as a task.   A task is a set of 32 resulting coefficients in a specific order. It tells a  warp how to calculate coefficients (the thread number one of the warp takes the first coefficient of the task and so on).
These tasks are reordered over the warps to balance the amount of work. An illustration of this algorithm is given on  figure \ref{algo_GPU}. 
The more coefficients the polynomials have, the more uniform the distribution of the tasks over the warps will be, and the more efficient the calculation.
We fix the maximum number of warps to eight (8 warps/256 threads), as far as possible, we assign at least two tasks per warp.
Thus, depending on  the number of coefficients a warp may calculate one task, whereas another warp may have several tasks to calculate.
When the determination of the list of the tasks is over,  we transfer  asynchronously  to the global memory the data of the polynomial and all the tasks.
This method does not  change the ordering of the data, ss for the CPU version, we have an AoS order.
But we have now informations how calculate the polynomial into a efficient way.
In terms of memory of the GPU, we allocate allocated only once, and reuse inner products after inner products., thus we save time.

@@

This algorithm gives good  performance on a GPU because the warps have approximatively the same amount of work,  and every thread in a warp will be responsible for all contributions to an output coefficient.
The contributions are not dispatched between the threads, thus we do not  introduces   synchronizations and leads to low performance. 
This  algorithm has the great advantage of working well with dense - triangular polynomials, and it could be extended to sparse polynomials, as structure of the polynomial is not primordial.
The main difficulty consists  of designing an efficient function GetIndices (\ref{ALGO}) which determines the contributions.

@@

The VLI library provides efficient solvers for the 128 to 512 bits integer associated with polynomials of 1 to 4 variables. We obtain a comfortable speed up as high as  60 compared 
to GMP for a specific range. The limited performances of GMP for small size were the main argument to develop the VLI library, although GMP excels in thousand bits arithmetic where our library can not operate. 
As future work, we would like to implement, new kinds of polynomials, and a better management  of  the memory to maintain the performance on GPU. And them, we would like explore new  SIMD architecture Xeon PHI where the carry bit is implemented.

>>>>

%\date{}                                           % Activate to display a given date or no dater
\begin{document}
\maketitle
%\section{}
%\subsection{} giamarchi
%\vspace{-0.65cm}
\begin{abstract}
The main purpose of the  Very Large Integer (VLI/C++/ASM) library is to perform large integer arithmetic and  inner products of polynomials. Based on meta-programming of assembly language,
and  hybrid (CPU/GPU) mode, it performs inner products of dense and triangular polynomials  of 1 to 4 variables  where the coefficients are represented  by large integers of 128 to 512 bits. 
In this paper, we present the basics of the inner products of polynomials, and  the implementation of the library for the CPU and the GPU.  All our algorithms are based on brute force attacks, because the problem is too small for a FFT approach. We save a factor up to 10 on CPU, and 60 on GPU compared to a GMP solution,
and attain 67 \% of the peak performance of the GPU.
\end{abstract}
\section{Introduction}
% TODO references LINBOX, NTL

The Very Large Integer library performs  standards  operations for large integer arithmetic and polynomial. This library also provides optimized implementations of inner products  of polynomials for GPU.
This library is composed of two distinct modules,  but enough closed  to work together by the mechanism of template instantiation. VLI is a fix size library from 128 to 512 bits.
The polynomial module  is limited for dense and triangular polynomials  up to 4 variables. The core of VLI is build under assembly language to provide the best performance.  Every arithmetic operations
are executed by a specific kernel generated specifically during the compilation by an assembly language generator. Moreover, VLI allows multithreading and hybrid  GPU usage for the inner products of polynomials.
Multithreading is build with OpenMP library whereas the GPU module utilizes CUDA. As on the CPU, PTX kernels are generated specifically during the compilation.

In large arithmetic, the standard library is GMP\footnote{GMP is the standard library for long arithmetic, gmplib.org}. GMP is polyvalent, nevertheless this polyvalence is attained by the usage of  dynamic memory.
It is well know dynamic allocation is time consuming, and it can become a bottle neck for small operations, which correspond to our working range.  
Alternatively, for polynomial arithmetic, a few libraries like  LINBOX or NTL  provide a solution but  they use GMP when needed. 
Although hardware accelerators were introduced  in HPC a few years ago, as the GPUs  by NVIDIA. Only a few libraries have been adapted 
A prominent example of such an adaption are the MAGMA libraries, which port the well known BLAS/LAPACK packages to GPU accelerators.
The polynomials libraries, however,  have not done the transition to these new hardwares yet.
During the last few years, the polynomial multiplication was successfully implemented on this new hardware
\cite{Govindaraju2008, Emeliyanenko:2009, Moreno2010}. All these works focus on large polynomials, using a FFT approach.

In this work, we are going to highlight the speedup given by a tiny ASM library without any dynamic memory,  suggest an alternative at GMP for small size integer. 
We also propose an efficient implementation of the inner products for  dense and triangular multivariable polynomials  on GPU with the latest CUDA PTX features.

@@

The library performs standards operations for polynomials (up to four variables) and optimized solvers for  the inner products of vectors of polynomials (presented in the next sections).
The polynomials are defined like  $e.g.$ our variables
\begin{eqnarray}
P_a(x,y,z,w) & = & \sum_{i=0}^n \sum_{j=0}^n  \sum_{k=0}^n \sum_{l=0}^n  a_{ijkl} x^i y^j z^k w^l , \,\,\,\,\,\, \text{or} \\
                   & = & \sum_{i=0}^n \sum_{j=0}^{(n-i)} \sum_{k=0}^{(n-i-j)} \sum_{l=0}^{(n-i-j-k)}  a_{ijkl} x^i y^j z^k w^l  .
\end{eqnarray}
where  $x,\,y,\,z$ and $w$ are symbolic variables,  $i,\,j,\,k$ and $l$  the indices, $a$  the coefficients, and $n$ the order of the polynomial.
The polynomials of the form  (1) and (2) are named dense and triangular respectively. Consider now two vectors of polynomials of the  same type with $m$ 
entries  $\boldsymbol{v}_1$ and  $\boldsymbol{v}_2$. The inner products  consist of the product 
of all entries  associated with a global reduction. 
 \begin{eqnarray}
 P_c = \sum_{i=0}^m \boldsymbol{v}_{1_i} \times  \boldsymbol{v}_{2_i}  \label{poly1} \label{InnerProduct}
\end{eqnarray}
where $\times$ indicates the multiplication between two polynomials. The multiplications use usual rules of polynomials multiplication. 
A final coefficient may be the sum of several coefficients (which we call contributions in the following passage), resulting from the sum of several multiplications. These coefficients are the cross terms. 
The cross terms are not a problem for the CPU version, as a polynomial multiplication is done by a single thread, but they are the bottleneck for the GPU version.
The number of coefficient for a dense and triangular polynomial is given by $(n+1)^{\kappa}$ and  $(\kappa+n-1)! / ( (\kappa-1)! n! )$ respectively, where $\kappa$ is the number of variables.

@@

\section{Very Large Integer library}

@@

We coded the kernels using the GNU inline assembly.
Generating by hand all these  assembly strings (kernels)  can be painful and error-prone\footnote{The library has near 10 000 lines of assembly in total for x86-64, Power64, and PTX}.
We designed a assembly generator. It creates all versions of a kernel during the compilation.   
It is built on Boost preprocessor package. We can thus create and manipulate the ASM strings during the compilation, 
and generate a specifically tuned kernel for every large integer. We give a small example of a long addition with the generator into the table \ref{ASMGENERATOR}.

@@

\subsection{Memory layout}

@@

We have chosen the SoA for the two following reasons :

@@

The disagreement  of this choice, the SoA order does not facilitate the usage of the SIMD instruction such as the AVX instructions of the Sandy Bridge x86-64 architecture.
Applying SIMD to the AoS for an addition/subtraction is a nonsense because there is no specific instruction to propagate carry bit between entries of the SIMD register.
It could be  apply  on multiplications for integer  larger than 1024-bit \cite{SIMD}, unfortunately, it does not correspond to our working range. Consequently, we do not 
use the multimedia instruction into the ASM kernels.

@@

The GPU algorithm of the inner product is divided into two parts. A preliminary step on the CPU executed only once,  where we prepare the execution blue print of the  contribution. And the second step the calculation itself on the GPU

@@

\item  First, we create a dummy resulting polynomial  represented by the indices and the contributions. They are sorted by descending order of the number of contributions
and grouped by 32 as a task.   A task is a set of 32 resulting coefficients in a specific order. It tells a  warp how to calculate coefficients (the thread number one of the warp takes the first coefficient of the task and so on).
These tasks are reordered over the warps to balance the amount of work. An illustration of this algorithm is given on  figure \ref{algo_GPU}. 
The more coefficients the polynomials have, the more uniform the distribution of the tasks over the warps will be, and the more efficient the calculation.
We fix the maximum number of warps to eight (8 warps/256 threads), as far as possible, we assign at least two tasks per warp.
Thus, depending on  the number of coefficients a warp may calculate one task, whereas another warp may have several tasks to calculate.
When the determination of the task lists is over,  we transfer  asynchronously  to the global memory the data of the polynomial and all the task lists.
This method does not  change the ordering of the data, as for the CPU version, we have an AoS order.
In terms of memory of the GPU, we allocate allocated only once, and reuse inner products after inner products, thus we do not wast time on successive reallocations of the memory.

@@

This algorithm is efficient on a GPU because the warps have approximatively the same amount of work,  and every thread in a warp will be responsible for all contributions to an output coefficient.
The contributions are not dispatched between the threads, thus we do not  introduces   synchronizations and leads to low performance. 
This  algorithm has the great advantage of working well with dense - triangular polynomials, and it could be extended to sparse polynomials, as structure of the polynomial is not primordial.
The main difficulty consists  of designing an efficient function \texttt{GetIndices} (table \ref{ALGO}) which determines the contributions.

@@

The VLI library provides efficient solvers for the 128 to 512 bits integer associated with polynomials of 1 to 4 variables. We obtain a comfortable speed up as high as  60 compared 
to GMP for a specific range. The limited performances of GMP for small size were the main argument to develop the VLI library, although GMP excels in thousand bits arithmetic where our library can not operate. 
As future work, we would like to implement, new kinds of polynomials, and a better management  of  the memory to maintain the performance on GPU. And then, we would like explore new  the SIMD architecture Xeon PHI 
where the carry bit is natively implemented.
